---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data
```{r loading, results="hide"}
########################################################################
## Comment on the report contents
########################################################################
## Since the assignment is not totally clear about what is the "report" and what
## is "the submission"
## (the .md or (.md + .Rmd) ?) I have made the choice to limit the amount of code 
## displayed in the .md file in order to keep it readable for a reader only
## interested in key results (I guess it would be the purpose in real business).
## 
## I included code in .md document only for the loading and preprocessing data section
## and for the strategy for imputing NA value.
## The rest of the .md is only plots and text to answer questions.
##
## Of course, all code is provided in the .Rmd file in order to be reproduced
## (most chuncks have echo=FALSE to be removed to get everything in the .md).
##
########################################################################
##
FIBIT_DATA<-read.csv("./activity.csv", sep=",", header= TRUE, na.strings=c("NA"),
                     colClasses=c("numeric","character","numeric"))
library(dplyr)
```

## What is mean total number of steps taken per day?
```{r step1, echo=FALSE}
##--- Calculate total number of steps for each day (61 days)
STEPS_PER_DAY<-summarize(group_by(FIBIT_DATA,date), sum(steps, na.rm = TRUE))

##--- Prepare Histogram
colnames(STEPS_PER_DAY) <- c("date","steps_per_day") #rename column in STEPS_PER_DAY
hist(STEPS_PER_DAY$steps_per_day, col = "blue", breaks = 15,
     main="-Total Steps per Day (Two Months) - mean (red) and median (green)-",
     xlab="-Number of Steps (One Day)-")
rug(STEPS_PER_DAY$steps_per_day)
##--- Mean and median values added to plot
abline(v=mean(STEPS_PER_DAY$steps_per_day), col = "red", lwd=4)
abline(v=median(STEPS_PER_DAY$steps_per_day), col = "green", lwd=4)

##--- mean and median to report
mean_steps_day<-as.integer(mean(STEPS_PER_DAY$steps_per_day))
median_steps_day<-as.integer(median(STEPS_PER_DAY$steps_per_day))
```

Mean number of steps per day is `r mean_steps_day`. 

Median number of steps per day is `r median_steps_day`.

## What is the average daily activity pattern?
```{r step2, echo=FALSE}
##--- Prepare Time Series Plot
STEPS_PER_INTERVAL<-summarize(group_by(FIBIT_DATA,interval), mean(steps, na.rm = TRUE))
colnames(STEPS_PER_INTERVAL) <- c("interval","mean_steps_per_interval") #rename columns
with(STEPS_PER_INTERVAL,{plot(interval, mean_steps_per_interval,
                              type="l", col="blue", lwd=2,
                              main="-Daily Average Steps per Interval-",
                              xlab="-Intervals-",
                              ylab="-Number of Steps (averaged for 2 Months)-")})

##-- Identify Daily Peak Activity
max_number_of_steps<-max(STEPS_PER_INTERVAL$mean_steps_per_interval)
peak_activity<-filter(STEPS_PER_INTERVAL,
                      mean_steps_per_interval == max_number_of_steps)
##--- Report Peak Activity 
peak_activity_hour<-substr(as.character(peak_activity$interval),1,1)
peak_activity_minute<-substr(as.character(peak_activity$interval),2,3)
```

The daily peak activity interval starts at `r peak_activity_hour`.
                                           `r peak_activity_minute` a.m. 

## Imputing missing values
```{r step3, echo=FALSE}
########################################################################
## Imputing Missing Values
########################################################################
##--- Calculate total number of NA
NA_Number<-nrow(filter(FIBIT_DATA, is.na(steps) == "TRUE"))
```

The total number of NA is : `r NA_Number`.

```{r step4}
################################################################
## Imputing Missing Values Strategy
################################################################
##--- Replace NA with mean value for same time interval
##--- new dataset FIBIT_DATA_NONA is created with "no NA" values.
FIBIT_DATA_NONA<-FIBIT_DATA
for (i in 1:nrow(FIBIT_DATA_NONA))
        {if(is.na(FIBIT_DATA_NONA[i,1])=="TRUE")
                {temp<-filter(STEPS_PER_INTERVAL,
                             interval == FIBIT_DATA_NONA[i,3])
                 FIBIT_DATA_NONA[i,1]<-as.integer(temp[1,2])
                }   
        }
```

```{r step5, echo=FALSE}
#--- Histogram on Updated FIBIT DATA (NO NA)
STEPS_PER_DAY_NONA<-summarize(group_by(FIBIT_DATA_NONA,date),
                              sum(steps, na.rm = TRUE))
colnames(STEPS_PER_DAY_NONA) <- c("date","steps_per_day") #rename columns
hist(STEPS_PER_DAY_NONA$steps_per_day, col = "blue", breaks = 15, lwd=2,
     main="-Total Steps per Day (Two Months) - mean (red) and median (green)-",
     xlab="-Number of Steps (One Day)-")
rug(STEPS_PER_DAY_NONA$steps_per_day)
abline(v=mean(STEPS_PER_DAY_NONA$steps_per_day), col = "red", lwd=4)
abline(v=median(STEPS_PER_DAY_NONA$steps_per_day), col = "green", lwd=4)
mean_steps_day_nona<-as.integer(mean(STEPS_PER_DAY_NONA$steps_per_day))
median_steps_day_nona<-as.integer(median(STEPS_PER_DAY_NONA$steps_per_day))
```

Mean number of steps per day is `r mean_steps_day_nona`. 

Median number of steps per day is `r median_steps_day_nona`.

Compared to the first part of the assignment the "left bar" has been reduced. Therefore, mean and median values have slightly increased and converged.

## Are there differences in activity patterns between weekdays and weekends?
```{r step6, echo=FALSE, fig.height=10}
#--- Add weekdays - weekend
FIBIT_DATA_NONA<-mutate(FIBIT_DATA_NONA, day_category = weekdays(as.POSIXct(date)))
for (i in 1:nrow(FIBIT_DATA_NONA))
        { if((FIBIT_DATA_NONA[i,4]=="Samedi")||(FIBIT_DATA_NONA[i,4]=="Dimanche"))
                {FIBIT_DATA_NONA[i,4]<-"weekend"} 
          else
                {FIBIT_DATA_NONA[i,4]<-"weekday"}
        }

#--- Prepare Plot Layout
par(mfrow = c(2,1))

#--- Weekdays (at the top)
STEPS_PER_INTERVAL_WD<-summarize(group_by(filter(FIBIT_DATA_NONA,
                                                 day_category=="weekday"),
                                          interval),
                                 mean(steps, na.rm = TRUE)
                                 )
colnames(STEPS_PER_INTERVAL_WD) <- c("interval","mean_steps_per_interval")
with(STEPS_PER_INTERVAL_WD,{plot(interval, mean_steps_per_interval,
                                 type="l", col="blue", lwd=2,
                                 ylim=c(0,250),
                                 main="-Daily Average Steps per Interval (Weekdays)-",
                                 xlab="-Intervals-",
                                 ylab="-Steps - Weekdays- Averaged for 2 Months-")
                           }
     )

#--- Weekends (Bottom plot)
STEPS_PER_INTERVAL_WE<-summarize(group_by(filter(FIBIT_DATA_NONA,
                                                 day_category=="weekend"),
                                          interval),
                                 mean(steps, na.rm = TRUE)
                                 )
colnames(STEPS_PER_INTERVAL_WE) <- c("interval","mean_steps_per_interval")
with(STEPS_PER_INTERVAL_WE,{plot(interval, mean_steps_per_interval, 
                                 type="l", col="green", lwd=2,
                                 ylim=c(0,250),
                                 main="-Daily Average Steps per Interval (Weekend)-",
                                 xlab="-Intervals-",
                                 ylab="-Steps - Weekends - Averaged for 2 Months-")
                            }
     )
```

The activity starts slightly later at WE. There is no "morning peak" but the activity level is higher (spcifically in the afternoon).
